Syntax and Semantics of Abstract Binding Trees

### 2015-11-13 paper.tex
Use todonotes
Clean up

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{agda} % this should come after ams*
\usepackage{catchfilebetweentags}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{ifdraft}
\usepackage{ifthen}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{scalerel}
\usepackage{setspace}
\usepackage{stmaryrd}
\usepackage{supertabular}
\usepackage{tikz-cd}
\usepackage{tikzpfeile}
\usepackage{url}

@@

Abstract binding trees (abts) are a generalization of abstract syntax trees
where operators may express variable binding structure as part of their
arities.  Originally formulated by Peter Aczel~\cite{aczel:1978}, unisorted
abts have been deployed successfully as the uniform syntactic framework for
several implementations of Constructive Type Theory, including Nuprl
\cite{constable:1986}, MetaPRL~\cite{hickey:2003} and JonPRL
\cite{jonprl:2015}.

@@

In parallel, M.\ Fiore and his collaborators have developed the categorical
semantics for several variations of second-order algebraic theories
\cite{fiore:1999, fiore:2005, fiore-hur:2010, fiore-mamoud:2010}; of these, the
simply sorted variants are equivalent to Harper's abstract binding trees
augmented with a notion of second-order variable (metavariable).

@@

\[
  \infer{
    \IsValence{\MkValence{\sigma_0,\dots,\sigma_m}{\tau_0,\dots,\tau_n}{\tau}}
  }{
    \IsSort{\tau} &
    \HypJ{\IsSort{\sigma_i}}{\Leq{i}{m}} &
    \HypJ{\IsSort{\tau_i}}{\Leq{i}{n}} &
  }
\]

@@

\[
  \infer{
    \IsArity{\MkArity{v_0,\ldots,v_n}{\tau}}
  }{
    \IsSort{\tau} &
    \HypJ{\IsValence{v_i}}{\Leq{i}{n}}
  }
\]

@@

By the Grothendieck construction
$\Iso{\Grothendieck{}{\Operators}}{\Comma{\mathbb{1}}{\Operators}}$ we have a category of
elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are paired
diagrams:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\Pair{\Upsilon}{a}}
      \arrow[r, "\Pair{\rho}{f}", swap]
    \& {\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
  \qquad\begin{tikzcd}[cramped, sep = small, ampersand replacement = \&]
    \& \mathbb{1}
      \arrow[dl, "\vartheta", swap]
      \arrow[dr, "\vartheta'"]
    \&
    \\
    {\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, "\Operators\Pair{\rho}{f}", swap]
    \&
    \& {\Operators\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]
We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\ra\SCtx\times\Discrete{\Arities}}$
in which they reside:
\begin{alignat*}{3}
  &\IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}&&\,\mathrel{\ {{:}}}\,&&\IMode{\SCtx\times\Discrete{\Arities}\ra\Sets}\\
  &\IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}&&\,=\,\ && %
    \OMode{%
      \MkSet{
        \vartheta\in\Pair{\Upsilon}{a}
        \mid
        {\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} =
        \Pair{\Upsilon}{a}
      }
    }
\end{alignat*}

@@

For instance, consider the $\lambda$-calculus with a single sort, $\SortExp$; we
give its signature $\Sigma_{\lambda}$ by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{lam}}{
    \MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}
These rules correspond to objects of $\Grothendieck{}{{\Operators}_{\lambda}}$ and we
collect their fibers into sets of operators:
\begin{alignat*}{5}
\IMode{{\pi_{{\Operators}_{\lambda}}}^{-1}}&\IMode{\Pair{\Upsilon}{{\MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}}}}\quad&&=
  &&\quad\OMode{\MkSet{\mathtt{lam}}}\\
\IMode{{\pi_{{\Operators}_{\lambda}}}^{-1}}&\IMode{\Pair{\Upsilon}{\MkArity{\MkValence{}{}{\SortExp},\MkValence{}{}{\SortExp}}{\SortExp}}}\quad&&=
  &&\quad\OMode{\MkSet{\mathtt{ap}}}\\
\IMode{{\pi_{{\Sigma}_{\lambda}}}^{-1}}&\triangleq\IMode{\bigcup_{\Pair{\Upsilon}{a}}{{\pi_{{\Operators}_{\lambda}}}^{-1}{\Pair{\Upsilon}{a}}}}\quad&&=
  &&\quad\OMode{\MkSet{\mathtt{lam},\ \mathtt{ap}}}\\
\end{alignat*}
So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):

\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{decl}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{\SortExp}{}{\SortExp}
    }{
      \SortExp
    }
  }\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{get}[u]}{
    \MkArity{}{\SortExp}
  }\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{set}[u]}{
    \MkArity{
      \MkValence{}{}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

@@

\begin{gather*}
  \infer{
    \IsMetaCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsMetaCtx{\Omega,\MV{m}:v}
  }{
    \IsMetaCtx{\Omega} &
    \IsValence{v} &
    \NotIn{\MV{m}}{\Dom\Omega}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsVarCtx{\Gamma,x:\tau}
  }{
    \IsVarCtx{\Gamma} &
    \IsSort{\tau} &
    \NotIn{x}{\Dom\Gamma}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:\tau}
  }{
    \IsVarCtx{\Upsilon} &
    \IsSort{\tau} &
    \NotIn{u}{\Dom\Upsilon}
  }
\end{gather*}

@@

\begin{gather*}
  \infer[\textit{var}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{\tau}
  }{
    \Lookup{\Gamma}{x}{\tau}
  }\\[6pt]
  %
  \infer[\textit{mvar}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{\tau}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{\sigma_0,\dots,\sigma_m}{\tau_0,\dots,\tau_n}{\tau}}\\
      \HypJ{\Lookup{\Upsilon}{u_i}{\sigma_i}}{\Leq{i}{m}}\\
      \HypJ{
        \IsAbt{\Omega}{\Upsilon}{\Gamma}{M_i}{\tau_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  %
  \infer[\textit{app}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{\tau}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{v_1,\dots,v_n}{\tau}
      }\\
      \HypJ{
        \IsBTm{\Omega}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  \infer[\textit{abs}]{
    \IsBTm{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsAbt{\Omega}{\Upsilon, \vec{u}:\vec{\sigma}}{\Gamma,\vec{x}:{\vec{\tau}}}{
      M
    }{
      \tau
    }
  }
\end{gather*}

Abts are identified up to $\alpha$-equivalence. Let $\FS{M}$ be the collection
of symbols free in $M$, and let $\FV{M}$ be the collection of variables free in
$M$. %% TODO: actually define these

@@

\begin{gather*}
  \infer{
    \IsRenaming{\varrho}{x}{x}
  }{
  }\\[6pt]
  \infer{
    \IsRenaming{\varrho}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{m}}{\vec{v}}{N_0,\dots,N_n}
    }
  }{
    \Match{\varrho(\vec{u})}{\vec{v}} &
    \HypJ{
      \IsRenaming{\varrho}{M_i}{N_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  \infer{
    \IsRenaming{\varrho}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \App{\vartheta'}{F_0,\dots,F_n}
    }
  }{
    \Match{\FLift{\varrho}{}(\vartheta)}{\vartheta'} &
    \HypJ{
      \IsRenaming{\varrho}{E_i}{F_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  \infer{
    \IsRenaming{\varrho}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{x}}{N}
    }
  }{
    \IsRenaming{\varrho\setminus\vec{u}}{M}{N}
  }
\end{gather*}

@@

%%% TODO: show that renaming preserves wellformedness

@@

\begin{gather*}
  \infer{
    \IsSubst{N}{x}{y}{N}
  }{
    \IsEq{x}{y}
  }\qquad
  \infer{
    \IsSubst{N}{x}{y}{y}
  }{
    \IsApart{x}{y}
  }\\[6pt]
  \infer{
    \IsSubst{N}{x}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{m}}{\vec{u}}{M'_0,\dots,M'_n}
    }
  }{
    \HypJ{
      \IsSubst{N}{x}{M_i}{M'_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  \infer{
    \IsSubst{N}{x}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \App{\vartheta}{F_0,\dots,F_n}
    }
  }{
    \HypJ{
      \IsSubst{N}{x}{E_i}{F_i}
    }{
      \Leq{i}{n}
    }
  }
\end{gather*}
\begin{gather*}
  \infer{
    \IsSubst{N}{x}{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{y}}{M'}
    }
  }{
    \NotIn{x}{\vec{y}} &
    \IsApart{\vec{u}}{\FS{N}} &
    \IsApart{\vec{y}}{\FV{N}} &
    \IsSubst{N}{x}{M}{M'}
  }\qquad
  \infer{
    \IsSubst{N}{x}{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }
  }{
    \Member{x}{\vec{y}} &
    \IsApart{\vec{u}}{\FS{N}} &
    \IsApart{\vec{y}}{\FV{N}}
  }
\end{gather*}

@@

\begin{gather*}
  \infer{
    \IsSubst{E}{\MV{m}}{x}{x}
  }{
  }\\[6pt]
  \infer{
    \IsSubst{E}{\MV{m}}{
      \MApp{\MV{n}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{n}}{\vec{u}}{N_0,\dots,N_n}
    }
  }{
    \IsApart{\MV{m}}{\MV{n}} &
    \HypJ{
      \IsSubst{E}{\MV{n}}{M_i}{N_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  \infer{
    \IsSubst{
      \MkBTm{\vec{u}}{\vec{x}}{N}
    }{\MV{m}}{
      \MApp{\MV{n}}{\vec{v}}{\vec{M}}
    }{
      N''
    }
  }{
    \IsEq{\MV{m}}{\MV{n}} &
    \IsSubst{\vec{M}}{\vec{x}}{N}{N'} &
    \IsRenaming{\MkSet{\vec{u}\mapsto\vec{v}}}{N'}{N''}
  }\\[6pt]
  \infer{
    \IsSubst{E}{\MV{m}}{
      \App{\vartheta}{F_0,\dots,F_n}
    }{
      \App{\vartheta}{F'_0,\dots,F'_n}
    }
  }{
    \HypJ{
      \IsSubst{E}{\MV{m}}{F_i}{F'_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  \infer{
    \IsSubst{E}{\MV{m}}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{x}}{N}
    }
  }{
    \IsApart{\vec{u}}{\FS{E}} &
    \IsApart{\vec{x}}{\FV{E}} &
    \IsSubst{E}{\MV{m}}{M}{N}
  }
\end{gather*}

@@

Then, $V$ is the unit to this tensor. We will say that an object
$\Of{P}{\widehat{\HypCat}^\Sorts}$ is a $\Sigma$-monoid in case it is equipped
with the following natural transformations where $\nu$ embeds variables into
$P$ and $\varsigma$ equips $P$ with an operation for simultaneous substitutions
of variables. Furthermore, $\nu$ and $\varsigma$ induce maps ${\nu}_{\Gamma}$ and
${{\varsigma}^{\tau}}_{\Upsilon\parallel\Gamma}$:
\vspace{-0.5em}
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{V}} \arrow[r, "\IMode\nu"]
    \& {\IMode{P}}
    \& {\IMode{P\odot P}}
    \arrow[l, "\IMode\varsigma", swap]
  \end{tikzcd}
  \qquad\begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{{V}^{[\Gamma]}}}
      \arrow[r, "\IMode{{\nu}_{\Gamma}}"]
    \& {\IMode{{P}^{[\Gamma]}}}
  \end{tikzcd}
  \qquad\begin{tikzcd}[cramped, sep = large, ampersand replacement = \&]
    {\IMode{{P_\tau^{\Yoneda{\Upsilon\parallel\Gamma}}\times S^{\{\Upsilon\}}\times P^{[\Gamma]}}}}
      \arrow[r, "\IMode{\varsigma^\tau_{\Upsilon\parallel\Gamma}}"]
    \& {\IMode{P_\tau}}
  \end{tikzcd}
\]

\begin{remark}
  A presheaf $P$ with a substitution monoidal structure $\Pair{\nu}{\varsigma}$
  also forms a \emph{relative monad}\footnote{Relative monads are skew-monoids
    in the skew-monoidal structure induced by a functor category
    $\mathbf{C}^\mathbf{J}$~\cite{altenkirch-chapman-uustalu:2015}. However,
    relative monads can also be understood without reference to monoids by
    viewing them as a generalization of \emph{extension systems} or
    \emph{Kleisli structures}. This latter presentation of monads, popular in
    computer science and logic, does not involve iteration of the monad
    functor.} on the functor $\Of{V}{\widehat{\HypCat}^\Sorts}$ of variables.
  Consequently, monoid multiplication $\varsigma$ can be replaced with extension
$\Of{\circledast}{\int_{\Tuple{\tau, (\Upsilon\parallel\Gamma), (\Upsilon'\parallel\Gamma')}}{{{P}_{\tau}(\Upsilon'\parallel\Gamma')}^{{V}_{\tau}(\Upsilon\parallel\Gamma)} \ra {{P}_{\tau}(\Upsilon'\parallel\Gamma')}^{{P}_{\tau}(\Upsilon\parallel\Gamma)}}}$.
\end{remark}

@@

%%% TODO: talk about context encodings in Abella

>>>>

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{agda} % this should come after ams*
\usepackage{catchfilebetweentags}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{ifdraft}
\usepackage{ifthen}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{scalerel}
\usepackage{setspace}
\usepackage{stmaryrd}
\usepackage{supertabular}
\usepackage{tikz-cd}
\usepackage{tikzpfeile}
\usepackage{todonotes}
\usepackage{url}

@@

Abstract binding trees (abts) are a generalization of abstract syntax trees
where operators may express variable binding structure as part of their arities.
Originally formulated by Peter Aczel~\cite{aczel:1978}, unisorted abts have been
deployed successfully as the uniform syntactic framework for several
implementations of Constructive Type Theory, including
Nuprl~\cite{constable:1986}, MetaPRL~\cite{hickey:2003} and
JonPRL~\cite{jonprl:2015}.

@@

In parallel, M.\ Fiore and his collaborators have developed the categorical
semantics for several variations of second-order algebraic
theories~\cite{fiore:1999, fiore:2005, fiore-hur:2010, fiore-mamoud:2010}; of
these, the simply sorted variants are equivalent to Harper's abstract binding
trees augmented with a notion of second-order variable (metavariable).

@@

\[
  \infer{
    \IsValence{
      \MkValence
        {\sigma_0,\dots,\sigma_m}
        {\tau_0,\dots,\tau_n}
        {\tau}
    }
  }{
    \IsSort{\tau}
&
    \HypJ{\IsSort{\sigma_i}}{\Leq{i}{m}}
&
    \HypJ{\IsSort{\tau_i}}{\Leq{i}{n}}
  }
\]

@@

\[
  \infer{
    \IsArity{
      \MkArity
        {v_0,\ldots,v_n}
        {\tau}
    }
  }{
    \IsSort{\tau}
&
    \HypJ
      {\IsValence{v_i}}
      {\Leq{i}{n}}
  }
\]

@@

By the Grothendieck construction
$\Iso{\Grothendieck{}{\Operators}}{\Comma{\mathbb{1}}{\Operators}}$ we have a
category of elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are paired
diagrams:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\Pair{\Upsilon}{a}}
      \arrow[r, swap, "\Pair{\rho}{f}"]
\&
    {\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\qquad
  \begin{tikzcd}[cramped, sep = small, ampersand replacement = \&]
    \null
\&
    \mathbb{1}
      \arrow[dl, swap, "\vartheta"]
      \arrow[dr, "\vartheta'"]
\&
    \null
\\
    {\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, swap, "\Operators\Pair{\rho}{f}"]
\&
    \null
\&
    {\Operators\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]
We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\ra\SCtx\times\Discrete{\Arities}}$
in which they reside:
\begin{alignat*}{3}
  \null
&
  \IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}
&&
  \,\mathrel{\ {{:}}}\,
&&
  \IMode{\SCtx\times\Discrete{\Arities}\ra\Sets}
\\
  \null
&
  \IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}
&&
  \,=\,\
&&
    \OMode{
      \MkSet{
        \vartheta\in\Pair{\Upsilon}{a}
        \mid{\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} =
        \Pair{\Upsilon}{a}
      }
    }
\end{alignat*}

@@

For instance, consider the $\lambda$-calculus with a single sort, $\SortExp$; we
give its signature $\Sigma_{\lambda}$ by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{lam}}{
    \MkArity{
      \MkValence{}{\SortExp}{\SortExp}
    }{\SortExp}
  }
\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}
These rules correspond to objects of $\Grothendieck{}{{\Operators}_{\lambda}}$
and we collect their fibers into sets of operators:
\begin{alignat*}{5}
  \IMode{{\pi_{{\Operators}_{\lambda}}}^{-1}}
&
  \IMode{
    \Pair{
      \Upsilon
    }{
      \MkArity{
        \MkValence{}{\SortExp}{\SortExp}
      }{\SortExp}
    }
  }
\quad
&&
  =
&&
\quad
  \OMode{\MkSet{\mathtt{lam}}}
\\
  \IMode{{\pi_{{\Operators}_{\lambda}}}^{-1}}
&
  \IMode{
    \Pair{
      \Upsilon
    }{
      \MkArity{
        \MkValence{}{}{\SortExp},
        \MkValence{}{}{\SortExp}
      }{\SortExp}
    }
  }
\quad
&&
  =
&&
\quad
  \OMode{\MkSet{\mathtt{ap}}}
\\
  \IMode{{\pi_{{\Sigma}_{\lambda}}}^{-1}}
&
  \triangleq\IMode{
    \bigcup_{\Pair{\Upsilon}{a}}{
      {\pi_{{\Operators}_{\lambda}}}^{-1}
      \Pair{\Upsilon}{a}
    }
  }
\quad
&&
  =
&&
\quad
  \OMode{
    \MkSet{\mathtt{lam},\ \mathtt{ap}}
  }
\\
\end{alignat*}
So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):

\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{decl}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{\SortExp}{}{\SortExp}
    }{
      \SortExp
    }
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{get}[u]}{
    \MkArity{}{\SortExp}
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{set}[u]}{
    \MkArity{
      \MkValence{}{}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

@@

\begin{gather*}
  \infer{
    \IsMetaCtx{\cdot}
  }{
  }
\qquad
  \infer{
    \IsMetaCtx{\Omega,\MV{m}:v}
  }{
    \IsMetaCtx{\Omega}
&
    \IsValence{v}
&
    \NotIn{\MV{m}}{\Dom\Omega}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }
\qquad
  \infer{
    \IsVarCtx{\Gamma,x:\tau}
  }{
    \IsVarCtx{\Gamma}
&
    \IsSort{\tau}
&
    \NotIn{x}{\Dom\Gamma}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }
\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:\tau}
  }{
    \IsVarCtx{\Upsilon}
&
    \IsSort{\tau}
&
    \NotIn{u}{\Dom\Upsilon}
  }
\end{gather*}

@@

\begin{gather*}
  \infer[\textit{var}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{\tau}
  }{
    \Lookup{\Gamma}{x}{\tau}
  }
\\[6pt]
  \infer[\textit{mvar}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{\tau}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{
        \MkValence{
          \sigma_0,\dots,\sigma_m
        }{
          \tau_0,\dots,\tau_n
        }{\tau}
      }
\\
      \HypJ{
        \Lookup{\Upsilon}{u_i}{\sigma_i}
      }{\Leq{i}{m}}
\\
      \HypJ{
        \IsAbt{\Omega}{\Upsilon}{\Gamma}{M_i}{\tau_i}
      }{\Leq{i}{n}}
    \end{array}
  }
\\[6pt]
  \infer[\textit{app}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{\tau}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{v_1,\dots,v_n}{\tau}
      }
\\
      \HypJ{
        \IsBTm{\Omega}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{\Leq{i}{n}}
    \end{array}
  }
\\[6pt]
  \infer[\textit{abs}]{
    \IsBTm{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsAbt{
      \Omega
    }{
      \Upsilon, \vec{u}:\vec{\sigma}
    }{
      \Gamma,\vec{x}:{\vec{\tau}}
    }{
      M
    }{\tau}
  }
\end{gather*}

Abts are identified up to $\alpha$-equivalence. Let $\FS{M}$ be the collection
of symbols free in $M$, and let $\FV{M}$ be the collection of variables free in
$M$. \todo[inline]{actually define these}

@@

\begin{gather*}
  \infer{
    \IsRenaming{\varrho}{x}{x}
  }{
  }
\\[6pt]
  \infer{
    \IsRenaming{\varrho}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{m}}{\vec{v}}{N_0,\dots,N_n}
    }
  }{
    \Match{\varrho(\vec{u})}{\vec{v}}
&
    \HypJ{
      \IsRenaming{\varrho}{M_i}{N_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer{
    \IsRenaming{\varrho}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \App{\vartheta'}{F_0,\dots,F_n}
    }
  }{
    \Match{\FLift{\varrho}{}(\vartheta)}{\vartheta'}
&
    \HypJ{
      \IsRenaming{\varrho}{E_i}{F_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer{
    \IsRenaming{\varrho}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{x}}{N}
    }
  }{
    \IsRenaming{\varrho\setminus\vec{u}}{M}{N}
  }
\end{gather*}

@@

\todo[inline]{show that renaming preserves wellformedness}

@@

\begin{gather*}
  \infer{
    \IsSubst{N}{x}{y}{N}
  }{
    \IsEq{x}{y}
  }
\qquad
  \infer{
    \IsSubst{N}{x}{y}{y}
  }{
    \IsApart{x}{y}
  }
\\[6pt]
  \infer{
    \IsSubst{N}{x}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{m}}{\vec{u}}{M'_0,\dots,M'_n}
    }
  }{
    \HypJ{
      \IsSubst{N}{x}{M_i}{M'_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer{
    \IsSubst{N}{x}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \App{\vartheta}{F_0,\dots,F_n}
    }
  }{
    \HypJ{
      \IsSubst{N}{x}{E_i}{F_i}
    }{
      \Leq{i}{n}
    }
  }
\end{gather*}
\begin{gather*}
  \infer{
    \IsSubst{N}{x}{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{y}}{M'}
    }
  }{
    \NotIn{x}{\vec{y}}
&
    \IsApart{\vec{u}}{\FS{N}}
&
    \IsApart{\vec{y}}{\FV{N}}
&
    \IsSubst{N}{x}{M}{M'}
  }
\qquad
  \infer{
    \IsSubst{N}{x}{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }
  }{
    \Member{x}{\vec{y}}
&
    \IsApart{\vec{u}}{\FS{N}}
&
    \IsApart{\vec{y}}{\FV{N}}
  }
\end{gather*}

@@

\begin{gather*}
  \infer{
    \IsSubst{E}{\MV{m}}{x}{x}
  }{
  }
\\[6pt]
  \infer{
    \IsSubst{E}{\MV{m}}{
      \MApp{\MV{n}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{n}}{\vec{u}}{N_0,\dots,N_n}
    }
  }{
    \IsApart{\MV{m}}{\MV{n}}
&
    \HypJ{
      \IsSubst{E}{\MV{n}}{M_i}{N_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer{
    \IsSubst{
      \MkBTm{\vec{u}}{\vec{x}}{N}
    }{\MV{m}}{
      \MApp{\MV{n}}{\vec{v}}{\vec{M}}
    }{
      N''
    }
  }{
    \IsEq{\MV{m}}{\MV{n}}
&
    \IsSubst{\vec{M}}{\vec{x}}{N}{N'}
&
    \IsRenaming{\MkSet{\vec{u}\mapsto\vec{v}}}{N'}{N''}
  }
\end{gather*}
\begin{gather*}
  \infer{
    \IsSubst{E}{\MV{m}}{
      \App{\vartheta}{F_0,\dots,F_n}
    }{
      \App{\vartheta}{F'_0,\dots,F'_n}
    }
  }{
    \HypJ{
      \IsSubst{E}{\MV{m}}{F_i}{F'_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer{
    \IsSubst{E}{\MV{m}}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{x}}{N}
    }
  }{
    \IsApart{\vec{u}}{\FS{E}}
&
    \IsApart{\vec{x}}{\FV{E}}
&
    \IsSubst{E}{\MV{m}}{M}{N}
  }
\end{gather*}

@@

Then, $V$ is the unit to this tensor. We will say that an object
$\Of{P}{\widehat{\HypCat}^\Sorts}$ is a $\Sigma$-monoid in case it is equipped
with the following natural transformations where $\nu$ embeds variables into $P$
and $\varsigma$ equips $P$ with an operation for simultaneous substitutions of
variables. Furthermore, $\nu$ and $\varsigma$ induce maps ${\nu}_{\Gamma}$ and
${{\varsigma}^{\tau}}_{\Upsilon\parallel\Gamma}$: \vspace{-0.5em}
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{V}} \arrow[r, "\IMode\nu"]
\&
    {\IMode{P}}
\&
    {\IMode{P\odot P}}
      \arrow[l, swap, "\IMode\varsigma"]
  \end{tikzcd}
\qquad
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{{V}^{[\Gamma]}}}
      \arrow[r, "\IMode{{\nu}_{\Gamma}}"]
\&
    {\IMode{{P}^{[\Gamma]}}}
  \end{tikzcd}
\qquad
  \begin{tikzcd}[cramped, sep = large, ampersand replacement = \&]
    {\IMode{{P_\tau^{\Yoneda{\Upsilon\parallel\Gamma}}\times S^{\{\Upsilon\}}\times P^{[\Gamma]}}}}
      \arrow[r, "\IMode{\varsigma^\tau_{\Upsilon\parallel\Gamma}}"]
\&
    {\IMode{P_\tau}}
  \end{tikzcd}
\]

@@

\todo[inline]{talk about context encodings in Abella}
